A Short and Mechanized Consistency Proof for Dependent Type Theory

### 2024-01-09 paper/paper.tex
Add auxiliary functions

%%
%% This is file `sample-acmsmall.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `acmsmall')
%%
%% IMPORTANT NOTICE:
%%
%% For the copyright see the source file.
%%
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-acmsmall.tex.
%%
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%%
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[nonacm]{acmart}
\usepackage{ottalt}
\usepackage{xspace}
\newcommand{\lang}{$\lambda^H$\xspace}
\inputott{rules}
%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

@@

The key technique that underlies our consistency proof is proof by
logical relation. In
Section~\ref{sec:spec}, we present the dependent type theory of
interest. In Section~\ref{sec:logreldep}, we give the definition
of the logical relation for the dependent type theory. Rather than
presenting the logical relation as an inductive-recursive definition,
we use the more elementary concept of a partial function to capture
the interpretation of types. The alternative representation requires us
to show that the set of equations indeed defines a partial function;
that is, for each input, there should always be a unique
output.
From the interpretation function, we can define the semantic
typing judgment for the set of lambda terms.
In Section~\ref{sec:logrelproof}, we prove the fundamental lemma,
which states that syntactic typing implies semantic typing. Once the
fundamental lemma is established, logical consistency follows as a
trivial corollary. In Section~\ref{sec:logrelmech}, we point out the
specifics related to the Coq mechanization of the proof described in earlier
section.
Finally, in Section~\ref{sec:relatedwork}, we give a short survey of
existing literature related to logical consistency about dependent
type theory.

@@

\mathit{Contexts}\\
[[G]]       & ::= & [[empty]]\ |\ [[G ++ A]] &  \\ \\
\mathit{Terms}\\
[[a]],[[b]],[[c]],[[t]],[[p]],[[A]],[[B]] & ::= & [[Set i]]\ |\ [[i]]\  |\ [[Void]]
                  & \mbox{universes, variables, empty type} \\
            & |   & [[Pi A B]]\ |\ [[\ A a]]\ |\ [[a b]]
                  & \mbox{function types, abstractions, applications} \\
            & |   & [[a ~ b : A ]]\ |\  [[refl]]\ |\ [[J t a b p]]
                  & \mbox{equality types, reflexivity proof, J eliminator} \\
            & |   & [[Bool]]\ |\  [[true]]\ |\  [[false]]\ |\  [[if a b0 b1]]
                  & \mbox{boolean type, true, false, if} \\ \\
\mathit{Substitution}\\
[[rho]] & \in & [[SNat -> STm]] &
\end{array}
\]
  \caption{Syntax of \lang}
  \label{fig:syntax}
\end{figure}

@@

The syntax of \lang can be found in Figure~\ref{fig:syntax}. As a
dependent type theory, terms and types are collapsed into the same
syntactic category. Dependent functions take the form $[[Pi A B]]$ and
we use the notation $[[A -> B]]$ when the output type $[[B]]$ is not
dependent on the input variable. $[[Set i]]$ represents the universe
type where $[[i]]$ ranges over the set of natural numbers.
 Finally,
\lang also includes an intensional identity type $[[a ~ b : A]]$ whose
proofs can be eliminated by the J-eliminator $[[J t a b p]]$, where
$[[p]]$ is an equality proof between $[[a]]$ and $[[b]]$, and $[[t]]$
is the term whose type is to be casted.

@@

Figure~\ref{fig:par} shows the definition of the parallel reduction
relation, which takes the form $[[a => b]]$. We use $[[a =>+ b]]$ to
represent its transitive closure, which in turn allows us to define
the coherence relation $[[a <=> b]]$. % The $*$ symbol is more commonly
% used for the transitive and reflexive closure of a relation. However,
% since we can show that parallel reduction is reflexive, the transitive
% closure of parallel reduction must also be reflexive.
We say that two terms $[[a]]$
and $[[b]]$ are coherent if they can eventually reduce to some common
term $[[c]]$ through parallel reduction. The symmetric notation of
coherence suggests that it is an equivalence relation, and in fact, we
will use coherence as the equational theory for our type conversion
rule. Now, we prove that coherence is indeed an equivalence relation.

First, we show that coherence is reflexive through the following
sequence of lemmas.
\begin{lemma}[Par Refl]
  \label{lemma:parrefl}
  For all terms $[[a]]$, $[[a => a]]$.
\end{lemma}

@@

\begin{lemma}[Coherence refl]
  \label{lemma:coherencerefl}
  For all terms $[[a]]$, $[[a <=> a]]$.
\end{lemma}

@@

The reflexivity of parallel reduction enables us to embed rules from
call-by-name semantics into parallel reduction, as the following lemma
shows.
\begin{lemma}[Par AbsCbn] For all $[[A]], [[a]],$ and $[[b]]$,
  \label{lemma:parabscbn}
  $[[(\ A a) b => b {a}]]$
\end{lemma}
\begin{proof}
  Immediate from Lemma~\ref{lemma:parrefl} and \rref{P-AppAbs}.
\end{proof}

Symmetry of coherence immediately falls from its definition.
\begin{lemma}[Coherence sym]
  \label{lemma:coherencesym}
  If $[[a <=> b]]$, then $[[b <=> a]]$.
\end{lemma}

Before we can prove transitivity, we need to show that parallel
reduction satisfies the diamond property.
\begin{lemma}[Par cong]
  \label{lemma:parcong}
  If $[[a0 => a1]]$ and $[[b0 => b1]]$, then $[[a0 { b0 } => a1 { b1 }]]$.
\end{lemma}
\begin{lemma}[Par diamond]
  \label{lemma:pardiamond}
  If $[[a => b0]]$ and $[[a => b1]]$, then there exists some term
  $[[c]]$ such that $[[b0 => c]]$ and $[[b1 => c]]$.
\end{lemma}
The congruence property (Lemma~\ref{lemma:parcong}) can be proven by
structural induction over the derivation of $[[a0 => a1]]$.
Likewise, Lemma~\ref{lemma:pardiamond} can be proven by structural induction
over the derivation of $[[a => b0]]$. The \rref{P-AppAbs} case requires
the use of Lemma~\ref{lemma:parcong}.

From Lemma~\ref{lemma:parcong} and \ref{lemma:parrefl}, we recover the
single substitution property as a simple corollary.
\begin{corollary}[Par subst]
  \label{lemma:parsubst}
  If $[[a0 => a1]]$, then $[[a0 {b} => a1 {b}]]$ for arbitrary $[[b]]$.
\end{corollary}

A relation that satisfies the
diamond property must also be confluent, meaning that its transitive
and reflexive closure is confluent.
\begin{lemma}[Par confluent]
  \label{lemma:parconfluent}
  If $[[a =>+ b0]]$ and $[[a =>+ b1]]$, then there exists some term
  $[[c]]$ such that $[[b0 =>+ c]]$ and $[[b1 =>+ c]]$.
\end{lemma}
While $[[a =>+ b]]$ is defined as the transitive closure of $[[a => b]]$,
it coincides with the transitive and reflexive closure of $[[a => b]]$ since $[[a
=> b]]$ is reflexive (Lemma~\ref{lemma:parrefl}).

The transitivity of the coherence relation follows as a corollary of
Lemma~\ref{lemma:parconfluent}.
\begin{lemma}[Coherence trans]
  \label{lemma:coherencetrans}
  If $[[a0 <=> a1]]$ and $[[a1 <=> a2]]$, then $[[a0 <=> a2]]$.
\end{lemma}
By the definition of coherence, there exists some term $[[b]]$ such that $[[a0 =>+ b0]]$,
$[[a1 =>+ b0]]$ and some term $[[b1]]$ such that $[[a1 =>+ b1]]$ and
$[[a2 =>+ b1]]$. By Lemma~\ref{lemma:parconfluent}, there exists some
term $[[c]]$ such that $[[b0 =>+ c]]$ and $[[b1 =>+ c]]$. It sufficies
to show that $[[a0 =>+ c]]$ and $[[a2 =>+ c]]$, both of which
trivially hold since the transitive closure $[[a =>+ b]]$ is transitive.
This concludes the proof that coherence is an equivalence relation.
\begin{lemma}[Coherence Equivalence]
  \label{lemma:coherenceequiv}
  The relation $[[a <=> b]]$ satisfies reflexivity, symmetry, and
  transitivity and therefore is an equivalence relation.
\end{lemma}

We note that a more conventional presentation of
\rref{T-Conv} would instead use full beta reduction as the base for
the definition of coherence. However, since full beta reduction
doesn't satisfy the diamond property, one typically needs parallel
reduction as an auxilliary definition to derive the confluence of full
beta reduction. The formulation of \lang is slightly more economical.

@@

For the rest of the section, we establish some important facts about
the logical relation that will be useful for proving fundamental lemma
in Section~\ref{sec:logrelproof}.

The relation $[[Interp I i A S]]$ satisfies the following inversion
principles.
\begin{lemma}[Inversion of the logical relation]
  \label{lemma:interpinv}\leavevmode
  \begin{enumerate}
  \item If $[[Interp I i Void S]]$, then $[[S = emptyset]]$.
  \item If $[[Interp I i Bool S]]$, then $[[S = { a | a =>+ true \/ a =>+ false   }]]$.
  \item If $[[Interp I i a ~ b : A S]]$, then $[[S = { p | p =>+ refl , a <=> b  }]]$.
  \item If $[[Interp I i Pi A B S0]]$, then there exists $[[S]],[[F]]$ such that:
    \begin{itemize}
    \item $[[Interp I i A S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
    \item $[[S0 = { b | forall a, (# a in S implies b a in F a #) }]]$
    \end{itemize}
  \item If $[[Interp I i Set j S]]$, then $[[j < i]]$ and $[[S = I j]]$.
  \end{enumerate}
\end{lemma}
\begin{proof}
  We only show the most involved function case; the rest follows a
  similar but simpler pattern. We start by
  inducting over the derivation of $[[Interp I i Pi A B S]]$. There
  are only two possible cases we need to consider.
  \begin{description}
  \item[\Rref{I-Pi}:] Immediate.
  \item[\Rref{I-Red}:] We are given that $[[Interp I i Pi A B S0]]$.
    We know that there exists some $[[A0]]$ and
    $[[B0]]$ such that $[[Pi A B => Pi A0 B0]]$ and $[[Interp I i Pi
    A0 B0 S0]]$. From the
    induction hypothesis, there exists $[[S]]$ and $[[F]]$ such that :
    \begin{itemize}
    \item $[[Interp I i A0 S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B0 { a } F a #)]]$
    \item $[[S0 = { b | forall a, (# a in S implies b a in F a #) }]]$
    \end{itemize}
    By inverting $[[Pi A B => Pi A0 B0]]$, we derive $[[A => A0]]$ and
    $[[B => B0]]$. By Lemma~\ref{lemma:parsubst}, we have $[[B {a} => B0 {a} ]]$ for all
    $[[a]]$. As a result, by \rref{I-Red}, the same $[[S]]$ and
    $[[F]]$ satisfies the following properties:
    \begin{itemize}
    \item $[[Interp I i A S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
    \end{itemize}
    The properties above are exactly the preconditions needed to apply
    \rref{I-Pi} for $[[Pi A B]]$ to finish off the proof.
  \end{description}
\end{proof}

@@

\begin{lemma}[Preservation of the logical relation]

@@

\begin{corollary}[Logical Relation for Coherent Types]

@@

\begin{lemma}[Logical relation is functional]

@@

Lemma~\ref{lemma:logreldeter} enables us to derive the following
alternative introduction and inversion principles for \rref{I-Pi}.
\begin{lemma}[Pi Intro Alt]
  \label{lemma:piintroalt}
  Suppose the following two statements hold:
  \begin{itemize}
  \item $[[Interp I i A S]]$
  \item $[[forall a, a in S implies (# exists S0 , Interp I i B {a} S0 #)]]$
  \end{itemize}
  Then we have $[[Interp I i Pi A B { b | forall a, (# a in S , forall
    S0, (# Interp I i B {a} S0,  b a in F a #) #) }]]$
\end{lemma}
\begin{proof}
  Let $[[F]]$ be the relation defined as follows:
  \[ (a,S) \in [[F]] \iff [[Interp I i B {a} S]] \]
  By the second bullet from the premise and
  Lemma~\ref{lemma:logreldeter}, $[[F]]$ is a function that is total
  on the set $[[S]]$. The conclusion then trivially follows from
  \rref{I-Pi}.
\end{proof}

\begin{lemma}[Pi Inv Alt]
  \label{lemma:piinvalt}
  Suppose $[[Interp I i Pi A B S]]$, then there exists some $[[S0]]$
  such that the following constraints hold:
  \begin{itemize}
  \item $[[Interp I i A S0]]$
  \item $[[forall a, (# a in S0 implies (# exists S1 , Interp I i B {a}
    S1 #) #)]]$
  \item $[[S = { b | forall a, (# a in S0 , forall
    S1, (# Interp I i B {a} S1,  b a in S1 #) #) }]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  Immedaite from Lemmas~\ref{lemma:interpinv} and \ref{lemma:logreldeter}.
\end{proof}

@@

The next lemma shows that our logical relation satisfies
cumulativity. That is, if a type has an interpretation at a lower
universe level, then we can obtain the same interpretation at a higher
universe level.
\begin{lemma}[Logical relation cumulativity]
  \label{lemma:logrelcumulativity}
  If $[[Interp I i0 A S]]$ and $[[i0 < i1]]$, then $[[Interp I i1 A S]]$.
\end{lemma}
\begin{proof}
  Trivial by structural induction over the derivation of $[[Interp I
  i0 A S]]$.
\end{proof}
Note that in the statement of Lemma~\ref{lemma:logrelcumulativity}, we
implicitly assume that $[[I]]$ is defined on the set of natural
numbers less than $[[i1]]$.

\begin{corollary}[Logical relation is functional with different levels]
  \label{lemma:logreldeterhet}
  If $[[Interp I i0 A S0]]$ and $[[Interp I i1 A S1]]$, then $[[S0 = S1]]$.
\end{corollary}
\begin{proof}
  Immediate from Lemma~\ref{lemma:logreldeter} and
  \ref{lemma:logrelcumulativity}.
\end{proof}

>>>>

%%
%% This is file `sample-acmsmall.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `acmsmall')
%%
%% IMPORTANT NOTICE:
%%
%% For the copyright see the source file.
%%
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-acmsmall.tex.
%%
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%%
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[nonacm]{acmart}
\usepackage{ottalt}
\usepackage{xspace}
\usepackage[para]{footmisc}
\newcommand{\dotv}[2]{\href{#1}{\texttt{#1}}{\texttt{:#2}}}
\newcommand{\lang}{$\lambda^H$\xspace}
\inputott{rules}
%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

@@

The key technique that underlies our consistency proof is proof by
logical relation. In
Section~\ref{sec:spec}, we present the dependent type theory of
interest. In Section~\ref{sec:logreldep}, we give the definition
of the logical relation for the dependent type theory. Rather than
presenting the logical relation as an inductive-recursive definition,
we use the more elementary concept of a partial function to capture
the interpretation of types. The alternative representation requires us
to show that the set of equations indeed defines a partial function;
that is, for each input, there should always be a unique
output.
From the interpretation function, we can define the semantic
typing judgment for the set of lambda terms.
In Section~\ref{sec:logrelproof}, we prove the fundamental theorem,
which states that syntactic typing implies semantic typing. Once the
fundamental theorem is established, logical consistency follows as a
trivial corollary. In Section~\ref{sec:logrelmech}, we point out the
specifics related to the Coq mechanization of the proof described in earlier
section.
Finally, in Section~\ref{sec:relatedwork}, we give a short survey of
existing literature related to logical consistency about dependent
type theory.

@@

\mathit{Contexts}\\
[[G]]       & ::= & [[empty]]\ |\ [[G ++ A]] &  \\ \\
\mathit{Terms}\\
[[a]],[[b]],[[c]],[[t]],[[p]],[[A]],[[B]] & ::= & [[Set i]]\ |\ [[i]]\  |\ [[Void]]
                  & \mbox{universes, variables, empty type} \\
            & |   & [[Pi A B]]\ |\ [[\ A a]]\ |\ [[a b]]
                  & \mbox{function types, abstractions, applications} \\
            & |   & [[a ~ b : A ]]\ |\  [[refl]]\ |\ [[J t a b p]]
                  & \mbox{equality types, reflexivity proof, J eliminator} \\
            & |   & [[Bool]]\ |\  [[true]]\ |\  [[false]]\ |\  [[if a b0 b1]]
                  & \mbox{boolean type, true, false, if} \\ \\
\mathit{Renaming}\\
[[xi]] & \in & [[SNat -> SNat]] & \\ \\
\mathit{Substitution}\\
[[rho]] & \in & [[SNat -> STm]] &
\end{array}
\]
  \caption{Syntax of \lang}
  \label{fig:syntax}
\end{figure}

\begin{figure}[h]
    \[
      \begin{array}{lll}
        \mathit{Curried\ Addition} \\
        add(n) & := & m \mapsto n + m \\ \\
        \mathit{Extension ([[xi]])} \\
        [[(xi .: j) 0]]  & := & 0 \\
        [[(xi .: j) Suc i]]  & := & [[xi j]] \\ \\

        \mathit{Extension ([[rho]])} \\
        [[(rho .: a) 0]]  & := & [[a]] \\
        [[(rho .: a) Suc i]]  & := & [[rho i]] \\ \\

        \mathit{Up ([[xi]])} \\
        [[up xi]] & := & (add(1) \circ [[xi]]) , 0 \\ \\

        \mathit{Renaming} \\
        [[i {xi}]] & := & [[xi i]] \\
        [[(Set i) {xi}]] & := & [[Set i]] \\
        [[Void {xi}]] & := & [[Void]]\\
        [[(Pi A B) {xi}]] & := & [[Pi A{xi} B{up xi}]] \\
        [[(\ A a) {xi}]] & := & [[\ A {xi} (#a {up xi}#)]] \\
        [[(a b) {xi}]] & := & [[a {xi} (# b {xi} #)]] \\
        [[Bool {xi}]] & := & [[Bool]] \\
        [[true {xi}]] & := & [[true]] \\
        [[false {xi}]] & := & [[false]] \\
        [[(if a b0 b1) {xi}]] & := & [[if a{xi} b0{xi} b1{xi}]] \\
        [[(a ~ b : A) {xi}]] & := & [[ a{xi} ~ b{xi} : A{xi}]] \\
        [[refl {xi}]] & := & [[refl]] \\
        [[(J t a b p ) {xi}]] & := & [[J t {xi} a {xi} b {xi} p{xi}]] \\ \\

        \mathit{Shift} \\
        [[up n A]] & := & A \langle add(n) \rangle \\ \\

        \mathit{Shift\ one} \\
        [[up1 A]] & := & \uparrow^1 [[A]] \\ \\

        \mathit{Lookup} \\
        [[(G ++ A) 0]] & := &  [[A]] \\
        [[(G ++ A) Suc i]] & := & [[G i]] \\ \\

        \mathit{Drop} \\
        [[drop 0 G]] & := & [[G]] \\
        [[drop Suc i (#G ++ A#)]] & := & [[drop i G]] \\ \\
      \end{array}
    \]
  \caption{Auxiliary Functions over Syntax}
  \label{fig:auxdef}
\end{figure}

@@

The syntax of \lang can be found in Figure~\ref{fig:syntax}. We use
the unscoped DeBruijn representation to leave less ambiguity for the
formulation of the system and simplify the reasoning about binders.
Figure~\ref{fig:auxdef} shows the auxiliary definitions over the term
syntax, including renaming, substitution, and operations over
the typing context or substitution.
As a
dependent type theory, terms and types are collapsed into the same
syntactic category. Dependent functions take the form $[[Pi A B]]$ and
we use the notation $[[A -> B]]$ when the output type $[[B]]$ is not
dependent on the input variable. $[[Set i]]$ represents the universe
type where $[[i]]$ ranges over the set of natural numbers.
 Finally,
\lang also includes an intensional identity type $[[a ~ b : A]]$ whose
proofs can be eliminated by the J-eliminator $[[J t a b p]]$, where
$[[p]]$ is an equality proof between $[[a]]$ and $[[b]]$, and $[[t]]$
is the term whose type is to be casted.

@@

Figure~\ref{fig:par} shows the definition of the parallel reduction
relation, which takes the form $[[a => b]]$. We use $[[a =>+ b]]$ to
represent its transitive closure, which in turn allows us to define
the coherence relation $[[a <=> b]]$.
We say that two terms $[[a]]$
and $[[b]]$ are coherent if they can eventually reduce to some common
term $[[c]]$ through parallel reduction. The symmetric notation of
coherence suggests that it is an equivalence relation, and in fact, we
will use coherence as the equational theory for our type conversion
rule. Now, we prove that coherence is indeed an equivalence relation.

First, we show that coherence is reflexive through the following
sequence of lemmas.
\begin{lemma}[Par Refl\footnote{\dotv{join.v}{Par\_refl}}]
  \label{lemma:parrefl}
  For all terms $[[a]]$, $[[a => a]]$.
\end{lemma}

@@

\begin{lemma}[Coherence refl\footnote{\dotv{join.v}{Coherent\_reflexive}}]
  \label{lemma:coherencerefl}
  For all terms $[[a]]$, $[[a <=> a]]$.
\end{lemma}

@@

The reflexivity of parallel reduction enables us to embed rules from
call-by-name semantics into parallel reduction, as the following lemma
shows.
\begin{lemma}[Par AbsCbn\footnote{\dotv{join.v}{P\_AppAbs\_cbn}}] For all $[[A]], [[a]],$ and $[[b]]$,
  \label{lemma:parabscbn}
  $[[(\ A a) b => b {a}]]$
\end{lemma}
\begin{proof}
  Immediate from Lemma~\ref{lemma:parrefl} and \rref{P-AppAbs}.
\end{proof}

Symmetry of coherence immediately falls from its definition.
\begin{lemma}[Coherence sym\footnote{\dotv{join.v}{Coherent\_symmetric}}]
  \label{lemma:coherencesym}
  If $[[a <=> b]]$, then $[[b <=> a]]$.
\end{lemma}

Before we can prove transitivity, we need to show that parallel
reduction satisfies the diamond property.
\begin{lemma}[Par cong\footnote{\dotv{join.v}{par\_cong}}]
  \label{lemma:parcong}
  If $[[a0 => a1]]$ and $[[b0 => b1]]$, then $[[a0 { b0 } => a1 { b1 }]]$.
\end{lemma}
\begin{lemma}[Par diamond\footnote{\dotv{join.v}{par\_confluent}}]
  \label{lemma:pardiamond}
  If $[[a => b0]]$ and $[[a => b1]]$, then there exists some term
  $[[c]]$ such that $[[b0 => c]]$ and $[[b1 => c]]$.
\end{lemma}
The congruence property (Lemma~\ref{lemma:parcong}) can be proven by
structural induction over the derivation of $[[a0 => a1]]$.
Likewise, Lemma~\ref{lemma:pardiamond} can be proven by structural induction
over the derivation of $[[a => b0]]$. The \rref{P-AppAbs} case requires
the use of Lemma~\ref{lemma:parcong}.

From Lemma~\ref{lemma:parcong} and \ref{lemma:parrefl}, we recover the
single substitution property as a simple corollary.
\begin{corollary}[Par subst\footnote{\dotv{join.v}{par\_subst}}]
  \label{lemma:parsubst}
  If $[[a0 => a1]]$, then $[[a0 {b} => a1 {b}]]$ for arbitrary $[[b]]$.
\end{corollary}

A relation that satisfies the
diamond property must also be confluent, meaning that its transitive
and reflexive closure is confluent.
\begin{lemma}[Par confluent\footnote{\dotv{join.v}{pars\_confluent}}]
  \label{lemma:parconfluent}
  If $[[a =>+ b0]]$ and $[[a =>+ b1]]$, then there exists some term
  $[[c]]$ such that $[[b0 =>+ c]]$ and $[[b1 =>+ c]]$.
\end{lemma}
While $[[a =>+ b]]$ is defined as the transitive closure of $[[a => b]]$,
it coincides with the transitive and reflexive closure of $[[a => b]]$ since $[[a
=> b]]$ is reflexive (Lemma~\ref{lemma:parrefl}).

The transitivity of the coherence relation follows as a corollary of
Lemma~\ref{lemma:parconfluent}.
\begin{lemma}[Coherence trans\footnote{\dotv{join.v}{Coherent\_transitive}}]
  \label{lemma:coherencetrans}
  If $[[a0 <=> a1]]$ and $[[a1 <=> a2]]$, then $[[a0 <=> a2]]$.
\end{lemma}
By the definition of coherence, there exists some term $[[b]]$ such that $[[a0 =>+ b0]]$,
$[[a1 =>+ b0]]$ and some term $[[b1]]$ such that $[[a1 =>+ b1]]$ and
$[[a2 =>+ b1]]$. By Lemma~\ref{lemma:parconfluent}, there exists some
term $[[c]]$ such that $[[b0 =>+ c]]$ and $[[b1 =>+ c]]$. It sufficies
to show that $[[a0 =>+ c]]$ and $[[a2 =>+ c]]$, both of which
trivially hold since the transitive closure $[[a =>+ b]]$ is transitive.
This concludes the proof that coherence is an equivalence relation.
\begin{lemma}[Coherence Equivalence]
  \label{lemma:coherenceequiv}
  The relation $[[a <=> b]]$ satisfies reflexivity, symmetry, and
  transitivity and therefore is an equivalence relation.
\end{lemma}

We note that a more conventional presentation of
\rref{T-Conv} would instead use full beta reduction as the base for
the definition of coherence. However, since full beta reduction
doesn't satisfy the diamond property, one typically needs parallel
reduction as an auxilliary definition to derive the confluence of full
beta reduction. Our formulation of \lang through parallel reduction
is slightly more economical.

@@

For the rest of the section, we establish some important facts about
the logical relation that will be useful for proving fundamental theorem
in Section~\ref{sec:logrelproof}.

The relation $[[Interp I i A S]]$ satisfies the following inversion
principles.
\begin{lemma}[Inversion of the logical relation]
  \label{lemma:interpinv}\leavevmode
  \begin{enumerate}
  \item If $[[Interp I i Void S]]$, then $[[S = emptyset]]$.\footnote{\dotv{semtyping.v}{InterpExt\_Void\_inv}}
  \item If $[[Interp I i Bool S]]$, then $[[S = { a | a =>+ true \/ a =>+ false   }]]$.\footnote{\dotv{semtyping.v}{InterpExt\_Bool\_inv}}
  \item If $[[Interp I i a ~ b : A S]]$, then $[[S = { p | p =>+ refl , a <=> b  }]]$.\footnote{\dotv{semtyping.v}{InterpExt\_Eq\_inv}}
  \item If $[[Interp I i Pi A B S0]]$, then there exists $[[S]],[[F]]$ such that:\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv}}
    \begin{itemize}
    \item $[[Interp I i A S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
    \item $[[S0 = { b | forall a, (# a in S implies b a in F a #) }]]$
    \end{itemize}
  \item If $[[Interp I i Set j S]]$, then $[[j < i]]$ and $[[S = I j]]$.\footnote{\dotv{semtyping.v}{InterpExt\_Univ\_inv}}
  \end{enumerate}
\end{lemma}
\begin{proof}
  We only show the most involved function case; the rest follows a
  similar but simpler pattern. We start by
  inducting over the derivation of $[[Interp I i Pi A B S]]$. There
  are only two possible cases we need to consider.
  \begin{description}
  \item[\Rref{I-Pi}:] Immediate.
  \item[\Rref{I-Red}:] We are given that $[[Interp I i Pi A B S0]]$.
    We know that there exists some $[[A0]]$ and
    $[[B0]]$ such that $[[Pi A B => Pi A0 B0]]$ and $[[Interp I i Pi
    A0 B0 S0]]$. From the
    induction hypothesis, there exists $[[S]]$ and $[[F]]$ such that :
    \begin{itemize}
    \item $[[Interp I i A0 S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B0 { a } F a #)]]$
    \item $[[S0 = { b | forall a, (# a in S implies b a in F a #) }]]$
    \end{itemize}
    By inverting $[[Pi A B => Pi A0 B0]]$, we derive $[[A => A0]]$ and
    $[[B => B0]]$. By Lemma~\ref{lemma:parsubst}, we have $[[B {a} => B0 {a} ]]$ for all
    $[[a]]$. As a result, by \rref{I-Red}, the same $[[S]]$ and
    $[[F]]$ satisfies the following properties:
    \begin{itemize}
    \item $[[Interp I i A S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
    \end{itemize}
    The properties above are exactly the preconditions needed to apply
    \rref{I-Pi} for $[[Pi A B]]$ to finish off the proof.
  \end{description}
\end{proof}

@@

\begin{lemma}[Preservation of the logical relation\footnote{\dotv{semtyping.v}{InterpExt\_preservation}}]

@@

\begin{corollary}[Logical Relation for Coherent Types\footnote{\dotv{semtyping.v}{InterpUnivN\_Coherent}}]

@@

\begin{lemma}[Logical relation is functional\footnote{\dotv{semtyping.v}{InterpExt\_deterministic}}]

@@

Lemma~\ref{lemma:logreldeter} enables us to derive the following
alternative introduction and inversion principles for \rref{I-Pi}.
\begin{lemma}[Pi Intro Alt\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_nopf}}]
  \label{lemma:piintroalt}
  Suppose the following two statements hold:
  \begin{itemize}
  \item $[[Interp I i A S]]$
  \item $[[forall a, a in S implies (# exists S0 , Interp I i B {a} S0 #)]]$
  \end{itemize}
  Then we have $[[Interp I i Pi A B { b | forall a, (# a in S , forall
    S0, (# Interp I i B {a} S0,  b a in F a #) #) }]]$
\end{lemma}
\begin{proof}
  Let $[[F]]$ be the relation defined as follows:
  \[ (a,S) \in [[F]] \iff [[Interp I i B {a} S]] \]
  By the second bullet from the premise and
  Lemma~\ref{lemma:logreldeter}, $[[F]]$ is a function that is total
  on the set $[[S]]$. The conclusion then trivially follows from
  \rref{I-Pi}.
\end{proof}

\begin{lemma}[Pi Inv Alt\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv\_nopf}}]
  \label{lemma:piinvalt}
  Suppose $[[Interp I i Pi A B S]]$, then there exists some $[[S0]]$
  such that the following constraints hold:
  \begin{itemize}
  \item $[[Interp I i A S0]]$
  \item $[[forall a, (# a in S0 implies (# exists S1 , Interp I i B {a}
    S1 #) #)]]$
  \item $[[S = { b | forall a, (# a in S0 , forall
    S1, (# Interp I i B {a} S1,  b a in S1 #) #) }]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  Immedaite from Lemmas~\ref{lemma:interpinv} and \ref{lemma:logreldeter}.
\end{proof}

@@

The next lemma shows that our logical relation satisfies
cumulativity. That is, if a type has an interpretation at a lower
universe level, then we can obtain the same interpretation at a higher
universe level.
\begin{lemma}[Logical relation cumulativity\footnote{\dotv{semtyping.v}{InterpExt\_cumulative}}]
  \label{lemma:logrelcumulativity}
  If $[[Interp I i0 A S]]$ and $[[i0 < i1]]$, then $[[Interp I i1 A S]]$.
\end{lemma}
\begin{proof}
  Trivial by structural induction over the derivation of $[[Interp I
  i0 A S]]$.
\end{proof}
Note that in the statement of Lemma~\ref{lemma:logrelcumulativity}, we
implicitly assume that $[[I]]$ is defined on the set of natural
numbers less than $[[i1]]$.

\begin{corollary}[Logical relation is functional with different levels\footnote{\dotv{semtyping.v}{InterpExt\_deterministic'}}]
  \label{lemma:logreldeterhet}
  If $[[Interp I i0 A S0]]$ and $[[Interp I i1 A S1]]$, then $[[S0 = S1]]$.
\end{corollary}
\begin{proof}
  Immediate from Lemma~\ref{lemma:logreldeter} and
  \ref{lemma:logrelcumulativity}.
\end{proof}
